%!TEX root = ../thesis.tex
%*******************************************************************************
%*********************************** First Chapter *****************************
%*******************************************************************************

\ifpdf
    \graphicspath{{Chapter1/Figs/Raster/}{Chapter1/Figs/PDF/}{Chapter1/Figs/}}
\else
    \graphicspath{{Chapter1/Figs/Vector/}{Chapter1/Figs/}}
\fi


%********************************** %First Section  **************************************
\chapter{Introduction}
% \nomenclature[z-PPC]{PPC}{Particles per cell}

Computed Tomography (CT) imaging is widely used in both medicine and industry for investigating the internal structure of objects in a non-destructive manner. It works by directing X-rays through a sample from a number of different angles, and measuring their attenuation at each angle. In the medical domain, we wish to minimise the radiation dose to the patient, so measurements are conducted at as few angles (views) as possible and with the lowest possible X-ray energy. This leads to artifacts from both Gaussian and Poisson noise, as well as from the limited number of views restricting the attainable information, making it difficult to accurately reconstruct the internal structure of the sample. Although industrial applications are often able to use higher energy X-rays and use many more views, there are still many cases where radiation dosage can damage the sample, or where only a limited number of views are possible due to movement \cite{}.

In this work, we aim to reproduce work by Ho et. al. \citet{ho_denoising_2020}, which introduces a method for improving the quality of very low-dose CT reconstructions by using a patch-based diffusion (PaDIS) model as a prior. The reproduction of CT and machine learning research is often difficult due to the relatively limited number of public CT datasets, concerns regarding patient privacy, and poorly engineered or maintained codebases. For this reason, we developed our reproduction natively within the LION framework \cite{lion}, which provides physically accurate CT projection operators, standardised datasets and metrics, and a well-engineered codebase for developing and testing machine learning methods for CT reconstruction. We have therefore oped to maximise the utility of this work and enable easy comparison to future work by using a default LION dataset and geometry for our reproduction.

As well as implementing the core PaDIS pipeline, we also re-implemented the patch-averaging \cite{ozdenizci_restoring_2023} and patch-stitching \cite{rumberger_how_2021} inspired variants discussed in the PaDIS paper. Furthermore, we also ran experiments with some of the methods already implemented in LION, choosing those which are similar to the comparison models used by the original PaDIS paper. 

The full methodology of our reproduction is described in \autoref{sec:method}, with the results presented in \autoref{sec:results}. These are then discussed in \autoref{sec:discussion}, along with comments on the reproducibility of the Hu et. al. paper, before conclusions are presented in \autoref{sec:conclusion}. A theoretical overview of CT reconstruction and the variance exploding noise conditional score model (NCSN) used by PaDIS now follows.

