\ifpdf     \graphicspath{{Sections/2_background_figs/Raster/}{Sections/2_background_figs/PDF/}{Sections/2_background_figs/}}
\else     \graphicspath{{Sections/2_background_figs/Vector/}{Sections/2_background_figs/}}
\fi

\chapter{Background} \label{chap:background}

Here, we present a summary of the theory and literature required to understand
the experiments presented in this work. We focus on CT reconstruction, diffusion
models and their use in solving inverse problems, and the PaDIS method. We then
summarise the literature relevant to PaDIS, including both its motivators and
derivatives. The formulation presented in Section~\ref{sec:basics} summarises
the key points presented by Kak and Stanley~\cite{Kak_Stanley}, and utilises
their notation.

\section{Basics of Computed Tomography} \label{sec:basics}

In X-ray CT, an object is illuminated by X-rays from multiple angles, with the
intensity of radiation transmitted through the object measured by detectors on
the opposite side of the object. Through this method, we aim to reconstruct a
spatial distribution of the linear attenuation coefficient, written $f(x,y)$ in
the two-dimensional case, and with units of inverse length. This gives the
probability per unit path length that an X-ray photon is removed from the beam
by absorption or scattering. Therefore, assuming X-rays are monoenergetic, the
transmitted intensity along a ray $L$ should obey the Beer-Lambert law,
\begin{equation} I=I_0 \exp\left(-\int_L f(x,y) \, ds\right), \end{equation}
where $I_0$ is the intensity of radiation incident on the object, $I$ is the
detected intensity, and $ds$ denotes integration along the ray. Taking
logarithms then gives the linearised projection \begin{equation}
P=-\log\left(\frac{I}{I_0}\right)=\int_L f(x,y)\, ds, \end{equation} meaning
log-normalised CT measurements are approximated by a line integral through the
attenuation image. This approach assumes that beams are narrow, monoenergetic
and non-diffracting, and neglects scattering, beam hardening and finite aperture
effects, all of which contribute in clinical CT. However, it provides a
tractable mathematical basis within which to consider reconstruction algorithms.

\section{Detector Geometries} \label{sec:detector-geometries}

\section{Line Integrals, Projections and Sinograms} \label{sec:
line-integrals-projections-sinograms}

For a fixed projection angle $\theta$, we can define a line in the image plane
by \begin{equation} x\cos \theta + y\sin \theta=t, \label{eq:line_t}
\end{equation} where $t$ is the perpendicular distance of the line from the
origin, as shown in Figure~\ref{fig:line_integral}. The integral over this line
can then be denoted as \begin{equation} P_\theta(t) = \int_{L_{(\theta,t)}} f(x,
y) \, ds, \end{equation} or equivalently in Dirac delta form, \begin{equation}
P_\theta(t)=\int_{-\infty}^{\infty} \int_{-\infty}^{\infty} f(x,y)\delta(x\cos
\theta+ y\sin \theta-t) \, dx\, dy. \end{equation} Notably, the projection
$P_\theta(t)$ is therefore also the Radon transform~\cite{radon} of $f(x,y)$ at
angle $\theta$. We collect a line or grid of these projections simultaneously,
each corresponding to the same $\theta$ alignment, but different values of $t$.
In a hypothetical parallel-beam scan, all of these $\theta$-aligned rays are
parallel, whereas in a realistic fan-beam or cone-beam scan, the angle refers to
the central projection angle, from which other rays diverge.

In practice, we sample these projection lines from $K$ different angular views,
$\theta_1, \theta_2, \dots, \theta_K$, and collect the projections within $N$
detector bins corresponding to $t_1, t_2, \dots, t_N$, forming an array of
measurements called a sinogram, denoted \begin{equation} Y_{j,
i}=P_{\theta_i}(t_j), \quad i=1,\dots, K, \quad j=1,\dots, N. \end{equation}
Here, each column is a line of projections at directed angle $\theta$, and each
row corresponds to one detector bin. By vectorising the image of $f(x,y)$ as
$\mathbf{x}\in \mathbb{R}^n$, and the sinogram as $\mathbf{y}\in \mathbb{R}^m$,
the discretised CT forward model can be written as \begin{equation}
\mathbf{y}=A\mathbf{x}+\boldsymbol{\eta}, \label{eq:forward-model}
\end{equation} where $A\in \mathbb{R}^{m \times n}$ is the discretised forward
projection operator, $\boldsymbol{\eta}$ encodes noise and modelling errors, and
$m=KN$ in 2D. We can then express this operator $A$ explicitly by denoting an
image pixel $\ell$ by $\Omega_\ell$, and using the pixel basis function
\begin{equation} \phi_\ell(x,y)= \begin{cases} 1, & (x,y)\in \Omega_\ell,\\ 0, &
\text{otherwise}, \end{cases} \end{equation} such that we can approximate the
attenuation field as \begin{equation} f(x,y)\approx \sum_{\ell=1}^n x_\ell
\phi_\ell(x,y), \end{equation} where $x_\ell$ is the value of pixel $\ell$.
Therefore, substituting into the line integral and comparing with
Equation~\eqref{eq:forward-model} gives \begin{align} P_{\theta_i}(t_j) &\approx
\sum_{\ell=1}^n x_\ell \int_{L_{(\theta_i,t_j)}} \phi_\ell(x,y)\,ds,\\ \implies
A_{(j,i),\ell} &=\int_{L_{(\theta_i, t_j)}}\phi_\ell(x,y)\, ds, \end{align} so
each row of $A$ corresponds to a single measured ray, with entries encoding the
path length of that ray within each image pixel. This matrix is generally
extremely large, and impractical to store or manipulate.

\section{The Fourier Slice Theorem} \label{sec:fourier-slice-theorem}

Using the perpendicular distance of a ray from the origin, Equation~\eqref{eq:
line_t}, and the distance along that ray given by \begin{equation} s=-x\sin
\theta+ y \cos\theta, \end{equation} give the projection at angle $\theta$ as
\begin{equation} P_\theta(t)= \int_{-\infty}^{\infty} f(t\cos\theta-s\sin\theta,
\;t\sin\theta+s\cos\theta)\,ds. \end{equation} Taking the Fourier transform of
this with respect to $t$ then gives \begin{equation} S_\theta(\omega) =
\int_{-\infty}^{\infty}\int_{-\infty}^{\infty} f(t\cos\theta-s\sin\theta,\;
t\sin\theta+s\cos\theta) e^{-2\pi i \omega t} \,ds\,dt, \end{equation} so
converting to the $(x,y)$ coordinate system gives \begin{equation}
S_\theta(\omega) = \int_{-\infty}^{\infty}\int_{-\infty}^{\infty} f(x,y)e^{-2\pi
i \omega(x\cos\theta+y\sin\theta)} \,dx\,dy = F(\omega \cos \theta, \omega \sin
\theta). \end{equation} Therefore, the Fourier transform of a projection gives
the two-dimensional Fourier transform of the object along the radial line at the
same angle.

\section{Dosage and Noise} \label{sec:dosage-and-noise}

The radiation dose imparted on a patient is directly proportional to the total
energy deposited in the tissue. For this reason, it is also directly
proportional to the energy of the photons used, the number of photons released
at each view, and the number of views acquired. Dosage can therefore be reduced
by decreasing any of these properties. However, these each decrease imaging
quality.

Measurement noise is intrinsically linked to the intensity of radiation used.
If $N_0$ photons are incident on the sample along a ray with projection value
$p$, then we expect to detect \begin{equation} \bar N = N_0 \exp(-p).
\end{equation} As this is a counting experiment, the noise due to counting
errors can be approximated as Poisson such that \begin{equation} N\sim
\operatorname{Poisson}(\bar N). \end{equation} The projection estimate is
\begin{equation} \hat p =-\log \left(\frac{N}{N_0}\right), \end{equation} so at
high photon counts where $\operatorname{Poisson}(\bar N)\approx \mathcal{N}(\bar
N, \bar N)$, Taylor expansion gives \begin{equation} \hat p \approx p+\eta,
\quad \eta \sim \mathcal N\left(0, \frac{1}{\bar N}\right). \end{equation} At
low photon counts, however, we may have $N=0$, in which case $\hat p=-\log 0$,
which is undefined. Furthermore, relative fluctuations are large, preventing
Taylor expansion, and in general, the noise after the logarithm becomes
non-Gaussian, signal-dependent, biased and possibly undefined~\cite{}, meaning
this domain is often better to model with the Poisson likelihood. Decreasing
intensity therefore increases the difficulty in modelling, and is not discussed
in detail in this work. However, interesting advances in using diffusion models
trained with these sorts of priors can be found at~\cite{}.

Decreasing the number of views also decreases the dosage, but in accordance
with the Fourier slice theorem, this also limits...