\ifpdf
    \graphicspath{{Sections/3_methods_figs/Raster/}{Sections/3_methods_figs/PDF/}{Sections/3_methods_figs/}}
\else
    \graphicspath{{Sections/3_methods_figs/Vector/}{Sections/3_methods_figs/}}
\fi

\chapter{Methods}
\label{sec:method}

\section{The Dataset}
In this work, we use the LIDC-IDRI dataset \cite{lidc} for all experiments. This dataset contains 1018 human chest CT scans as raw CT DICOM \cite{DICOM} images from 1010 patients, with each scan containing many axial slices, and some slices containing cancerous nodules. Before training or reconstruction, we convert these into a processed slice dataset, saving each slice as a $512\times 512$ NumPy array, along with a nodule mask. All but the 512-resolution experiments then downsample these to $256\times256$ resolution during dataloading. Although the dataset contains multiple scans for some patients, the LION processing pipeline keeps only one per patient, preventing accidental data leakage.

Division into train, validation and test sets is done by sorted patient ID in an 80:10:10 ratio, ensuring that slices from no patient are present across sets. The default PaDIS training regime then selects at most $4$ slices per patient, targeting a balance between nodule-containing and nodule-free slices. This is, however, constrained by the availability of annotations for each patient, so some contribute fewer than $4$ slices. The detailed breakdown is presented in \autoref{tab:lidc_split_sizes}.

\begin{table}[ht] 
    \centering 
    \begin{tabular}{lccc} 
        \hline \textbf{Split} & \textbf{Patient IDs} & \textbf{Images in 4-slice regime} & \textbf{Images in all-slice regime} \\
        \hline 
        Train & 809 & 2713 & 189725 \\ 
        Validation & 101 & 328 & 27426 \\ 
        Test & 102 & 326 & 25719 \\ 
        \hline Total & 1012 & 3367 & 242870 \\ 
        \hline 
    \end{tabular}
    \caption[Processed LIDC-IDRI split sizes.]{Processed LIDC-IDRI split sizes used in this work. Note that the dataset contains two unused patient IDs.}
    \label{tab:lidc_split_sizes} 
\end{table}

\section{Experimental Geometry}
In this work, all experiments use a 2D fan-beam CT geometry implemented using LION's CT operators, which are themselves based on the tomosipo implementations \cite{}. We treat each LIDC-IDRI slice independently with a $300\,\mathrm{mm}\times300\,\mathrm{mm}$ in-plane field of view. By default, experiments reconstruct a $256\times 256$ grid, whilst the explicitly labelled native resolution experiments reconstruct a $512\times 512$ grid. The detector contains $900$ bins over a $900\, \mathrm{mm}$ detector width, with a $575\,\mathrm{mm}$ source-to-origin distance and a $1050\,\mathrm{mm}$ source-to-detector distance. In this work, we do not add measurement noise so that we can isolate the effects of sparse sampling and limited angular coverage from photon-count statistics.

Fan-beam geometry tests whether the PaDIS conclusions transfer beyond the original parallel-beam setup and reproduces the finite-source divergence used by clinical scanners \cite{}. It does not make the experiment clinically realistic. The operator remains two-dimensional and noise-free, and omits scatter, beam hardening, motion and cone-beam effects.

\section{Model Training}
\label{sec:methods-model-training}

In this work, we train three classes of model to aid reconstruction. These are patch-based diffusion models, whole-image diffusion models and PnP denoisers.

\subsection{Diffusion Model Training}
The learned diffusion priors were trained following the variance-exploding regime described in \secref{sec:score_based_sdes}, and their denoisers can therefore be converted to scores and used for both unconditional generation and solving inverse problems as explained in \secref{subsec:diffusion_models_inverse_problems} and \secref{subsec:padis_approach}.

All diffusion denoisers were trained using the U-Net-like, noise-conditional NCSN++ model, $F_{\boldsymbol{\phi}}$, developed by Song et al. \cite{song...}, which allows multi-scale inputs. EDM preconditioning then converts the raw network output into the denoised prediction $D_{\boldsymbol{\phi}}$, as described in \secref{subsec:edm_preconditioning}. This network passes the noise level to each residual block, and the default models take three input channels: the noisy image and the $[-1,1]$ normalised x- and y-coordinates of each pixel. The positional encoding ablation instead takes only the noisy image channel. Throughout, we use dropout level 0.05, a base channel size of $128$, $4$ residual blocks per level, and attention resolution $16$. These models always use four resolution levels, with the patch-based models using $128$, $256$, $256$, and $256$ channels at each level, and the whole-image models using $128$, $256$, $256$, and $512$ channels for greater coarse-level capacity.

All diffusion models were trained using the Adam optimiser \cite{kingma}, with a learning rate of $2 \times 10^{-4}$, and the EDM preconditioning described in \secref{subsec:edm_preconditioning} with $\sigma_{\mathrm{data}}=0.5$. The whole-image models were trained for $24$ hours, and patch-based models trained for $12$ hours on an RTX PRO 6000 GPU. In both cases, the learning rate ramped up over the first 10 million seen training examples or patches. During diffusion model training, noise scales were sampled from an EDM-style log-normal distribution, and truncated to $\sigma\in[0.002,40]$, with $\log\sigma\sim\mathcal{N}(-1.2,1.2^2)$.

Using these conventions, whole-image diffusion denoisers were trained to denoise entire $256\times 256$ CT slices. The default $256\times256$ patch-based denoisers were trained with patch sizes $56\times56$, $32\times32$, and $16\times16$, where selection probabilities at each training step are 0.5, 0.3, and 0.2 respectively. The $512\times512$ model instead uses patch sizes $64\times64$, $32\times32$, and $16\times16$ with the same probabilities. Ablation models were also trained using all available processed slices, different patch sizes, and without positional encoding channels. The patch size schedules used for these ablations are detailed in \autoref{tab:patch_size_schedules}.

The diffusion denoisers were evaluated on the validation set throughout training to ensure overfitting did not occur. However, for the final six hours of each run, the validation frequency and intensity were significantly increased. The exponential-moving-average checkpoint corresponding to the minimum mean EDM validation loss during this period, referred to as \texttt{min\_intense\_val}, was then used for all reconstruction and generation experiments.

\begin{table}[ht] 
    \centering 
    \begin{tabular}{lccc} 
        \hline 
        \textbf{Ablation} & \textbf{Sampled patch sizes} & \textbf{Sampling probabilities} & \textbf{Padding width} \\ 
        \hline 
        $P=8$ & $8$ & $1.0$ & $8$ \\ 
        $P=16$ & $8,\ 16$ & $0.3,\ 0.7$ & $16$ \\ 
        $P=32$ & $8,\ 16,\ 32$ & $0.2,\ 0.3,\ 0.5$ & $32$ \\ 
        $P=56$ & $16,\ 32,\ 56$ & $0.2,\ 0.3,\ 0.5$ & $24$ \\ 
        $P=96$ & $32,\ 64,\ 96$ & $0.2,\ 0.3,\ 0.5$ & $32$ \\ 
        \hline 
    \end{tabular} 
    \caption[PaDIS patch-size training schedules.]{Patch-size schedules used for the PaDIS patch-size ablations. Each training step samples one patch size according to the listed probabilities. The $P=56$ row is the default PaDIS patch schedule used for the main 256-resolution patch model.}
    \label{tab:patch_size_schedules}
\end{table}

\subsection{PnP Denoiser Training}
Two PnP denoisers were trained using the existing PnP-ADMM \cite{chan_plug-and-play_2016} LION implementation, which uses a DRUNet model \cite{zhang_plug-and-play_2021} as the denoiser. At each training step, Gaussian noise is sampled uniformly from $\sigma \in[0,0.05]$ and added to the clean image, from which the model then learns to recover the clean target. 

Each model was trained using the same four-slice training split as the diffusion models, the Adam optimiser, a learning rate of $1\times 10^{-4}$ for $100$ epochs, batch size $8$, and MSE denoising loss. One model was trained with only the noisy image as input, and another was trained with the noise level additionally provided as a constant image channel. Both used $64$ internal channels, $4$ blocks per level, bias-free convolutions, and leaky-ReLU activations. Validation was performed after each epoch using the same Gaussian denoising objective, and for each model, the checkpoint corresponding to the lowest validation loss was used for inference experiments. During reconstruction, the noise-conditioned model is provided with a fixed noise level of $0.03$.

\section{Reconstructor Implementation}
\label{sec:methods-reconstructor-implementation}

The paper and code do not specify every method consistently (\secref{sec:discussion-reproducibility}) \cite{}. We therefore created two reproduction tracks: Paper-Style follows the written descriptions, while Public-Compatible follows the released code as part of the combined study specification. LION-physics is a separate extension using CT operator normalisation to improve stability and transfer between geometries. Each reproduction track exists only where the original study gives enough detail, whereas LION-physics was implemented for all methods. \autoref{tab:methods_by_implementation} summarises the available combinations, with differences detailed in \secref{subsec:public-compatible-implementation} and \secref{subsec:lion-physics-implementations}. Paper-Style uses the methodologies in \secref{sec:reconstruction_methods}.

\begin{table}[ht]
  \centering
  \caption[Reconstruction methods by implementation track.]{Reconstruction methods implemented in each track. A star indicates an additional modification or numerical stabilisation described below.}
  \label{tab:methods_by_implementation}
  \begin{tabular}{lccc}
  \hline
  \textbf{Method} & \textbf{Paper-style} & \textbf{Public-compatible} & \textbf{LION-physics} \\
  \hline
  Baseline FDK              & --             & --             & \(\checkmark\) \\
  Chambolle--Pock TV        & --             & --             & \(\checkmark^*\) \\
  PnP-ADMM                  & --             & --             & \(\checkmark\) \\
  Whole-image diffusion     & \(\checkmark\) & --             & \(\checkmark\) \\
  Langevin                  & \(\checkmark\) & \(\checkmark^*\) & \(\checkmark\) \\
  Predictor-corrector       & \(\checkmark^*\) & \(\checkmark^*\) & \(\checkmark\) \\
  VE-DDNM                   & \(\checkmark^*\) & \(\checkmark^*\) & \(\checkmark^*\) \\
  Patch averaging           & --             & \(\checkmark^*\) & \(\checkmark\) \\
  Patch stitching           & --             & \(\checkmark^*\) & \(\checkmark\) \\
  PaDIS-DPS                 & \(\checkmark\) & \(\checkmark^*\) & \(\checkmark\) \\
  \hline
  \end{tabular}
\end{table}

All three tracks use the same LIDC-IDRI images, trained denoisers, and LION fan-beam measurements. Paper-Style therefore refers to the equations and sampling choices described in the paper, rather than exact reproduction of its AAPM and parallel-beam experiments. LION-physics is evaluated as an extension and is not used to claim direct reproduction of the original algorithms. Reconstruction hyperparameters were re-tuned for the LION operators.

Throughout, an outer step means moving to the next noise level, while an inner step repeats the sampler update at the current noise level. Unless stated otherwise, diffusion reconstructions use 100 geometrically spaced outer levels from \(\sigma_{\max}=10\) to \(\sigma_{\min}=0.002\), with 10 inner updates at each level. The 8-view experiments instead use \(\sigma_{\min}=0.003\). The schedule ablation changes only the spacing of the outer levels from geometric to EDM-style. Predictor-corrector is an exception because each adjacent pair receives one predictor and one corrector update. Paper-style and LION-physics VE-DDNM instead use 1000 outer levels with one update per level, while Public-compatible VE-DDNM retains the released code's 100 outer levels with 10 inner updates.

The analytic baseline applies LION's FDK reconstruction. Hu et al. used ADMM-TV as their iterative baseline \cite{}. We retain its isotropic TV objective but use LION's CP implementation from \secref{subsec:iterative_reconstruction_denoising_priors} \cite{}. Validation selected \(\lambda=0.001\) and 1000 iterations. Outputs were clipped to \([0,1]\) without non-negativity. We therefore report CP rather than ADMM-TV. PnP-ADMM remains separate.

\subsection{Public-Compatible Implementations}
\label{subsec:public-compatible-implementation}
These methods were implemented to match the publicly available codebase, which differs from the paper descriptions in several important ways \cite{}. Its DPS implementation differentiates the norm of the residual and begins from FDK, whilst predictor-corrector evaluates the corrector at the current noise level and reuses its patch layout. VE-DDNM uses $100$ noise levels with $10$ updates at each, rather than $1000$ separate levels.

These rules, clipping and patch randomisation are retained with fixed scales for LION geometry. Final experiments use the paper's geometric schedule, with the public EDM-style schedule as an ablation. To validate this translation, we adapted the released code in the companion \texttt{PaDIS} directory to use the same LIDC-IDRI slices, checkpoint and geometry. Across three matched DPS reconstructions, target means were $33.12$ versus $33.16\,\mathrm{dB}$ PSNR and $0.836$ versus $0.834$ SSIM, while cross-implementation SSIM was $0.996$. The remaining difference was within sampler variation. Other public helpers were checked through patch layouts and intermediate states.

\subsection{LION-physics Implementations}
\label{subsec:lion-physics-implementations}
LION-physics extends the original work by removing these fixed scaling values. Following \autoref{eq:scaled_data_consistency_update}, the squared-residual gradient is instead divided by $L_F=\lVert F\rVert_2^2$, where $F$ maps the normalised image to its sinogram and its norm is estimated using power iteration. This keeps the meaning of the data-consistency coefficient more consistent between geometries, but does not guarantee convergence of the complete sampler.

These implementations otherwise use the paper's geometric schedule. Predictor-corrector uses the paper-style noise-level and patch-layout choices, whilst VE-DDNM clips its FDK pseudoinverse terms to prevent unstable values. Operator normalisation is only applied to diffusion data updates.

Patch averaging and stitching use fixed layouts with an eight-pixel overlap. Averaging takes the mean of all predictions covering each pixel, whilst stitching allows later patches to replace earlier values in overlaps. Both use DPS, but unlike PaDIS they do not repeatedly redraw shifted patch partitions.

\subsection{Hyperparameter Tuning}
Reconstruction hyperparameters were selected using only the validation set. Finite runs were ranked by mean PSNR, with differences below $0.1\,\mathrm{dB}$ treated as ties, then by SSIM within $0.01$, and finally by MAE. Where possible, one setting was selected across the 8- and 20-view experiments. Higher-view experiments use this setting when no dedicated tuning is available, whilst the 512-resolution experiment is tuned separately. The selected values are stored in LION and listed in the project README.

\section{Experiments}

The primary comparison uses noise-free sinograms from 8 and 20 uniformly spaced views over \(360^\circ\) at \(256\times256\) resolution. These experiments contain the method and implementation combinations listed in \autoref{tab:methods_by_implementation} and form the basis of \autoref{tab:main-quality}. Paper-style and Public-compatible test the central PaDIS claim under two readings of the combined paper-and-code specification, while LION-physics evaluates the operator-normalised extension on the same images, priors and fan-beam measurements.

Three extensions test transfer beyond that comparison. Selected high-performing and diagnostic methods are evaluated with 60 views over \(360^\circ\), and with 20 views restricted to a \(120^\circ\) arc. The latter retains the legacy experiment key \texttt{ct\_fanbeam\_180}, but is a limited-angle stress test rather than the paper's 180-view experiment. Native-resolution experiments use 60 full-angle views on a \(512\times512\) grid and compare FDK, CP-TV, public-compatible PaDIS-DPS and LION-physics PaDIS-DPS. No whole-image model was trained at this resolution.

All quantitative ablations use the 20-view full-angle problem. The patch-size, dataset-size and positional-encoding experiments revisit questions studied by Hu et al. \cite{}. The initialisation and schedule comparisons additionally resolve differences between the written algorithms and released code. The patch-size experiment compares maximum patch sizes \(P\in\{8,16,32,56,96\}\) with whole-image diffusion. The dataset-size experiment compares the four-slice-per-patient training regime with all processed training slices for both patch and whole-image priors. The position experiment crosses presence of coordinate channels with FDK or Gaussian initialisation, while the schedule ablation compares the two outer-level spacings defined above. Sampler comparisons apply Langevin, predictor-corrector, VE-DDNM and VE-DPS to patch and, where implemented, whole-image priors.

Unconditional generation presents four samples from each of four priors. These are whole-image diffusion, fixed patch stitching, fixed patch averaging and randomly shifted PaDIS. Each uses 300 geometric outer levels with one inner update per level. These outputs are assessed only as demonstrations of global anatomical coherence and boundary behaviour.

Standard reconstruction jobs use seed 33 and the same first 25 test images. Fixed patch-averaging and patch-stitching rows, unconditional generation, and all \(512\times512\) reconstructions use four images or samples because of their substantially greater cost. Results from these four-image subsets are identified explicitly and are not treated as equally precise estimates.

\section{Metrics}

All quantitative image metrics are computed against the held-out target after normalising both images to \([0,1]\). The report presents mean absolute error (MAE), peak signal-to-noise ratio (PSNR), and structural similarity (SSIM) \cite{}. For \(N\) image pixels,
\begin{align}
\mathrm{MAE}(\hat{\mathbf{x}},\mathbf{x})
&=\frac{1}{N}\lVert\hat{\mathbf{x}}-\mathbf{x}\rVert_1,\\
\mathrm{PSNR}(\hat{\mathbf{x}},\mathbf{x})
&=10\log_{10}\left(\frac{1}{\mathrm{MSE}(\hat{\mathbf{x}},\mathbf{x})}\right),
\end{align}
where the peak value is one. SSIM measures local agreement in luminance, contrast and structure. Higher PSNR and SSIM and lower MAE indicate closer agreement, but none establishes that clinically important features have been preserved.

Measurement consistency is reported separately as the relative sinogram residual
\begin{equation}
\mathrm{RSR}(\hat{\mathbf{x}})
=
\frac{\lVert A\hat{\mathbf{x}}-\mathbf{y}\rVert_2}
{\lVert\mathbf{y}\rVert_2}.
\end{equation}
Low RSR shows that a reconstruction explains the simulated measurements. It does not by itself imply similarity to the target because sparse and limited-angle operators have large poorly constrained subspaces.

Tables report the image-level mean and asymmetric central 95\% confidence interval from 2000 deterministic bootstrap resamples using seed 33 \cite{}. The bootstrap samples images, not training runs, so it captures variation across the evaluated slices but not uncertainty from retraining. For figures, normalised images are clipped to \([0,1]\), and ``Normalised Intensity'' is abbreviated to NI. HU panels use the mapping in \autoref{eq:hounsfield_units}. Each row is clipped to the 15th and 95th percentiles of non-background ground-truth pixels, defined by normalised intensity above 0.02. The same limits are applied to every reconstruction and the target in that row, preventing method-dependent contrast adjustment. Colour bars report the applied limits. The 50-mm scale bars use the configured 300-mm reconstruction field of view rather than patient-native DICOM spacing. Unconditional generations have no paired target and are assessed qualitatively.
\section{Compute and Reproducibility}
Training, tuning and reconstruction used a Google-hosted NVIDIA RTX PRO 6000 GPU and seed 33. The pipeline was independently developed in LION, with the released repository used only to check the Public-compatible track. Checkpoints were selected before testing. Weights \& Biases records training, and artefacts are retained on persistent cloud storage. Reproduction scripts and instructions are in \texttt{scripts/paper\_scripts/PaDIS}.

% - Implemented in LION - Produced three implementation tracks: Explicit paper
% implementation, public-repo matching setup, and one which uses correct operator
% normalisations to increase cross-problem reliability, as described in
% \ref{subsec:lion-physics}. - Created LION-physics implementations for all
% experiments, but only paper and public-repo matching implementations for those
% which were actually described in sufficient detail for reproduction in those
% respective soruces. This is summarised in \autoref{tab:implemented-methods}.
